Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-26 trunk/paper/drinking.tex

\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Gabriella
  Rustici$^2$ and Midori A. Harris$^3$ and Georgios V. Gkoutos$^{1}$}

@@

While the PATO framework is now successfully being applied to the
integration of phenotypes in which

@@

Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities. Examples of
classes that may have processes as instances include {\em Drinking},
{\em Triathlon} or {\em Apoptosis}.  Figure \ref{fig:onto} illustrates
these basic categories of being.

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process.
We can further distinguish different {\em roles} of participation in a
process. For example, a runner may participate in a {\em Triathlon}
process as a {\em Runner (role)}, while another person can participate
in the same process but in a different role (such as {\em Referee}).

>>>>

\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
  A. Harris$^3$ and ... and Gabriella Rustici$^2$ and Georgios
  V. Gkoutos$^{1}$}

@@

While the PATO framework is now successfully being applied to
semantically integrate phenotypes across species, the diversity and
complexity of phenotypes in which biological processes and functions
are impaired continues to limit the interoperability between phenotype
ontologies. Biological processes and functions are linked to the
physiology which the components of biological systems exhibit and

@@

Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities
\cite{Burek2006}. Examples of classes that may have processes as
instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
Figure \ref{fig:onto} illustrates these basic categories of being.

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process
\cite{Herre2006, Herre2010}.  We can further distinguish different
{\em roles} of participation in a process \cite{Loebe2007}. For
example, a runner may participate in a {\em Triathlon} process as a
{\em Runner (role)}, while another person can participate in the same
process but in a different role (such as {\em Referee}).
